Embedding Intuitionistic into Classical Logic

### 2022-07-14 CSL/lipics-v2021-EIICL.tex
Typos

Constructive mathematics refers to a flavour of mathematics in which existence of an object can only be established by explicit construction, as opposed to classical mathematics where existence can be shown implicitly, e.g. by assuming non-existence and deriving a contradiction. Besides the philosophical issues that one may take with classical mathematics, most prominently advocated by Brouwer~\cite{brouwer1907over} and Bishop~\cite{bishop1967foundations}, there is a particular motivation when looking from the position of computer science in that intuitionistic proofs correspond directly to computer programs, which is expressed in the Curry-Howard-Isomorphism~\cite{howard1980formulae}. The formalism usually associated with constructive mathematics is intuitionistic logic, which essentially differentiates itself from classical logic by the facts that the law of excluded middle $A\vee\neg A$ and double negation shift $\forall x\neg\neg P(x)\to\neg\neg\forall xP(x)$ are not valid.

However comparing benchmarks for classical and intuitionistic automated theorem proving, e.g. TPTP~\cite{casc} and ILTP~\cite{iltp}, we see that classical provers currently have a far better success rate. Furthermore all the most popular and used first-order provers~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover} utilize classical logic. This difference can be partially explained by fundamental differences between the logics. First of all in the propositional case determining intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic} whereas the analogous decision problem for classical logic is only \verb+coNP+-complete~\cite{cook1971complexity}. A further advantage of classical logic is the existence of calculi that are particularly suitable for automation like superposition~\cite{bachmair2001resolution} which rely on the existence of convenient normal forms such as CNF and the duality between validity and satisfiability, i.e. to show validity they show unsatisfiability of the negation. While some kind of normal form would of course obtainable in the intuitionistic case, crucially the duality does not hold, e.g. $\neg(A\vee \neg A)$ is not satisfiable while $A\vee\neg A$ is not a theorem. Therefore most dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient, tammet1996resolution} use the naive inverse method, i.e. directly constructing a cut-free proof, enhanced by search strategies such as focussing and polarization. This strategy generally leads to a much more complex search and is difficult to implement efficiently. Finally in contrast to intuitionistic provers a tremendous amount of work has been put into optimizing provers for classical logic, this is particularly true for the propositional case, i.e. SAT-solvers.

@@

In addition to the translation of formulas we give an effective translation of counter-models, i.e. for each intuitionistic counter-model for $\varphi$ we give a classical counter-model for $\varphi^\#$ and vice versa. In fact the existence of counter-models is a key notion. For classical logic it just corresponds to the satisfiability of the negation. However for intuitionistic logic it forms a proper dual to validity, while satisfiability of the negation is not necessary for invalidity and would therefore not form a proper dual. Transforming and reducing counter-models is what ultimately enables our translation.

@@

We suggest a new approach for obtaining a constructive proof using a classical prover:
\begin{itemize}
	\item Performing our translation on the input formula $\varphi$ to obtain $\varphi^\#$.
	\item Checking the classical validity of $\varphi^\#$ with the prover resulting in a proof or counter-model.
	\item Translating the proof/counter-model of $\varphi^\#$ to a (intuitionistic) proof/counter-model of $\varphi$.
\end{itemize}

>>>>

Constructive mathematics refers to a flavour of mathematics in which existence of an object can only be established by explicit construction, as opposed to classical mathematics where existence can be shown implicitly, e.g. by assuming non-existence and deriving a contradiction. Besides the philosophical issues that one may take with classical mathematics, most prominently advocated by Brouwer~\cite{brouwer1907over} and Bishop~\cite{bishop1967foundations}, there is a particular motivation for studying constructive mathematics from the perspective of computer science in that intuitionistic proofs correspond directly to computer programs, which is expressed in the Curry-Howard-Isomorphism~\cite{howard1980formulae}. The formalism usually associated with constructive mathematics is intuitionistic logic, which essentially differentiates itself from classical logic by the fact that the law of excluded middle $A\vee\neg A$ and the double negation shift $\forall x\neg\neg P(x)\to\neg\neg\forall xP(x)$ are not valid.

Comparing benchmarks for classical and intuitionistic automated theorem proving, e.g. TPTP~\cite{casc} and ILTP~\cite{iltp}, we see that classical provers currently have a far better success rate. Furthermore all the most popular first-order provers~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover} utilize classical logic. This difference can be partially explained by fundamental differences between the logics. First of all determining intuitionistic validity is computationally harder, i.e. in the propositional case intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic} whereas classical validity is \verb+coNP+-complete~\cite{cook1971complexity}. A further advantage of classical logic is the existence of calculi that are particularly suitable for automation like superposition~\cite{bachmair2001resolution} which rely on the existence of convenient normal forms such as CNF and the duality between validity and satisfiability, i.e. to show validity they show unsatisfiability of the negation. While some kind of normal form is of course be obtainable for intuitionistic logic, albeit more complex, crucially the duality does not hold, e.g. $\neg(A\vee \neg A)$ is not satisfiable while $A\vee\neg A$ is also not a theorem. Therefore most dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient, tammet1996resolution} use the naive inverse method, i.e. directly constructing a cut-free proof from the theorem by applying the inverse of rules from some proof calculus, enhanced by search strategies such as focussing and polarization. This strategy generally leads to a much more complex search and is therefore difficult to implement efficiently. Finally in contrast to intuitionistic provers a tremendous amount of work has been put into optimizing provers for classical logic, this is particularly true for the propositional case, i.e. SAT-solvers.

@@

In addition to the translation of formulas we give an effective translation of counter-models, i.e. for each intuitionistic counter-model for $\varphi$ we give a classical counter-model for $\varphi^\#$ and vice versa. In fact the existence of counter-models is a key notion. For classical logic it just corresponds to the satisfiability of the negation but for intuitionistic logic it forms a proper dual to validity, while satisfiability of the negation is not necessary for invalidity and would therefore not form a proper dual. Transforming and reducing counter-models to a normal form is also what ultimately enables our translation.

@@

Our work suggests a new approach for obtaining a constructive proof using a classical prover, that is
\begin{itemize}
	\item performing our translation on the input formula $\varphi$ to obtain $\varphi^\#$,
	\item checking the classical validity of $\varphi^\#$ with the prover resulting in a proof or counter-model,
	\item translating the proof/counter-model of $\varphi^\#$ to a (intuitionistic) proof/counter-model of $\varphi$.
\end{itemize}
